Hydra Battles and Cie

### 2021-08-09 doc/part-hydras.tex
converting chapter 4 to alectryon; add a link to Gaia

@@

\begin{Coqsrc}
Compute alpha_0.
\end{Coqsrc}

\begin{Coqanswer}
  = ocons omega 0 (ocons (FS 2) 4 (FS 1))
     : T1
\end{Coqanswer}

@@

\begin{Coqsrc}
Inductive ppT1 : Set :=
    P_fin : nat -> ppT1
  | P_add : ppT1 -> ppT1 -> ppT1
  | P_mult : ppT1 -> nat -> ppT1
  | P_exp : ppT1 -> ppT1 -> ppT1
  | P_omega : ppT1

Notation _omega := P_omega.

@@

\begin{Coqsrc}
Compute pp alpha_0.
\end{Coqsrc}

\begin{Coqanswer}
     = (_omega ^ _omega + _omega ^ 3 * 5 + 2)%pT1
     : ppT1
\end{Coqanswer}

@@

In order to compare two terms of type \texttt{T1}, we define a recursive function \texttt{compare} that maps two ordinals $\alpha$ and $\beta$ to a value of type \texttt{comparison}. This type is defined in \coq's standard library 
\texttt{Init.Datatypes} and
contains three constructors:  \texttt{Lt} (less than), \texttt{Eq} (equal), and
\texttt{Gt} (greater than).

@@

\begin{Coqsrc}
Fixpoint compare (alpha beta:T1):comparison :=
  match alpha, beta with
    zero, zero => Eq
  | zero, ocons a' n' b' => Lt
  | _   , zero => Gt
  | (ocons a n b),(ocons a' n' b') =>
      (match compare a a' with 
       | Lt => Lt
       | Gt => Gt
       | Eq => (match Nat.compare n n' with
                | Eq => compare b b'
                | comp => comp
                end)
       end)
  end.
\end{Coqsrc}

It is now easy to define the boolean predicate \texttt{lt\_b $\alpha$ $\beta$}: 
`` $\alpha$ is strictly less than $\beta$ ''. By coercion to sort \texttt{Prop} we define also the predicate \texttt{lt}.

\begin{Coqsrc}
Definition lt_b alpha beta : bool :=
  match compare alpha beta with
      Lt => true
    | _ => false
  end.

Definition lt alpha beta : Prop := lt_b alpha beta.
\end{Coqsrc}

@@

\begin{Coqsrc}
Example E1 : lt (ocons omega 56 zero) (tower 3).
Proof. reflexivity. Qed.

Example E2 : ~ lt (tower 3) (tower 3).
Proof.  discriminate.  Qed.
\end{Coqsrc}

The following lemmas establish relations between \texttt{compare}, 
the predicate \texttt{lt} and Leibniz equality \texttt{eq}.

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.Epsilon0.T1.html\#compare_refl}{Epsilon0.T1}}

\begin{Coqsrc}
Lemma compare_refl : forall alpha, compare alpha alpha =  Eq.
\end{Coqsrc}

\begin{Coqsrc}
Lemma compare_reflect : forall alpha beta,
    match compare alpha beta with
    |   Lt => lt alpha  beta
    |   Eq => alpha = beta
    |   Gt => lt beta  alpha
    end.
\end{Coqsrc}

We prove also that the relation \texttt{lt} is a strict total order.

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.Epsilon0.T1.html\#lt_irrefl}{Epsilon0.T1}}

\begin{Coqsrc}
Theorem lt_irrefl (alpha: T1):  ~ lt alpha alpha.

Theorem lt_trans (alpha beta gamma : T1) :
  lt alpha  beta -> lt beta gamma -> lt alpha gamma.

Definition lt_eq_lt_dec  :
   forall alpha beta : T1, 
          {lt alpha  beta} + {alpha = beta} + {lt beta alpha}.
\end{Coqsrc}

Note that the order \texttt{lt} is not reflected 
in the structure (size and/or height) of the terms of \texttt{T1}.

\begin{Coqsrc}
Example Ex0:
  lt (ocons (phi0 (phi0 omega)) 2
            (ocons (phi0 10) 33
                   (ocons (phi0 9) 63 zero)))
     (ocons  (phi0 (phi0 omega)) 2 (phi0 (phi0 11))).
Proof.
  reflexivity. 
Qed.
\end{Coqsrc}

@@

\begin{Coqbad}
Example bad_term  : T1 := ocons 1 1 (ocons omega 2 zero).
\end{Coqbad}

@@

\begin{Coqsrc}
Fixpoint nf_b (alpha : T1) : bool :=
  match alpha with
    | zero => true
    | ocons a n zero => nf_b a
    | ocons a n ((ocons a' n' b') as b) =>
      (nf_b a && nf_b b && lt_b a' a)%bool
  end.

Definition nf alpha: Prop := nf_b alpha.
\end{Coqsrc}

\begin{Coqsrc}
 Compute nf_b alpha_0.
\end{Coqsrc}

\begin{Coqsrc}
 Compute nf_b bad_term.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma plus_is_zero alpha beta :
  nf alpha -> nf beta ->
  alpha + beta  = zero -> alpha = zero /\  beta = zero.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition restrict {A:Type}(E: Ensemble A)(R: relation A) :=
    fun a b => E a /\ R a b /\ E b.
 \end{Coqsrc}

@@

\begin{Coqsrc}
Definition LT := restrict nf lt.
Infix "t1<" := LT : t1_scope.

Definition LE := restrict nf le.
Infix "t1<=" := LE : t1_scope.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma LT_one : forall alpha, alpha t1< one -> alpha = zero.
\end{Coqsrc}

@@

>>>>

\label{sect:comparable-def}

@@

\input{movies/snippets/T1/alpha0Compute}

@@

\input{movies/snippets/T1/ppT1Def}

@@

\input{movies/snippets/T1/ppAlpha0}

@@

In order to compare two terms of type \texttt{T1}, we define a recursive function \texttt{compare} that maps two ordinal tems $\alpha$ and $\beta$ to a value of type \texttt{comparison}. This type is defined in \coq's standard library 
\texttt{Init.Datatypes} and
contains three constructors:  \texttt{Lt} (less than), \texttt{Eq} (equal), and
\texttt{Gt} (greater than).

@@

\input{movies/snippets/T1/compareDef}

@@

\input{movies/snippets/T1/ltExamples}

%%% TODO : link to Comparable and StrictOrder classes

Links between the function \texttt{compare} and the relations
\texttt{lt} and \texttt{eq} are established through the following lemmas (~\vref{sect:comparable-def}).
\vspace{4pt}

\input{movies/snippets/T1/Instances}

% \vspace{4pt}
% \emph{From Module~\href{../theories/html/hydras.Epsilon0.T1.html\#compare_refl}{Epsilon0.T1}}

% \begin{Coqsrc}
% Lemma compare_refl : forall alpha, compare alpha alpha =  Eq.
% \end{Coqsrc}

% \begin{Coqsrc}
% Lemma compare_reflect : forall alpha beta,
%     match compare alpha beta with
%     |   Lt => lt alpha  beta
%     |   Eq => alpha = beta
%     |   Gt => lt beta  alpha
%     end.
% \end{Coqsrc}

% We prove also that the relation \texttt{lt} is a strict total order.

% \vspace{4pt}
% \emph{From Module~\href{../theories/html/hydras.Epsilon0.T1.html\#lt_irrefl}{Epsilon0.T1}}

% \begin{Coqsrc}
% Theorem lt_irrefl (alpha: T1):  ~ lt alpha alpha.

% Theorem lt_trans (alpha beta gamma : T1) :
%   lt alpha  beta -> lt beta gamma -> lt alpha gamma.

% Definition lt_eq_lt_dec  :
%    forall alpha beta : T1, 
%           {lt alpha  beta} + {alpha = beta} + {lt beta alpha}.
% \end{Coqsrc}

% \vspace{4pt}

% Please note that the order \texttt{lt} is not reflected 
% in the structure (size and/or height) of the terms of \texttt{T1}.

% \begin{Coqsrc}
% Example Ex0:
%   lt (ocons (phi0 (phi0 omega)) 2
%             (ocons (phi0 10) 33
%                    (ocons (phi0 9) 63 zero)))
%      (ocons  (phi0 (phi0 omega)) 2 (phi0 (phi0 11))).
% Proof.
%   reflexivity. 
% Qed.
% \end{Coqsrc}

@@

\input{movies/snippets/T1/badTerm}

@@

\input{movies/snippets/T1/nfDef}

\input{movies/snippets/T1/nfAlpha0}

\input{movies/snippets/T1/nfBadTerm}

@@

\input{movies/snippets/T1/plusIsZero}

@@

\input{movies/snippets/Restriction/restrictionDef}

@@

\input{movies/snippets/T1/LTDef}

@@

\input{movies/snippets/T1/LTOnea}

\begin{Coqanswer}
1 subgoal (ID 738)

  alpha1 : T1
  n : nat
  alpha2 : T1
  H1 : nf (ocons alpha1 n alpha2)
  H2 : lt (ocons alpha1 n alpha2) one
  ============================
  ocons alpha1 n alpha2 = zero
\end{Coqanswer}

\input{movies/snippets/T1/LTOneb}

@@

The Gaia project ~\url{https://github.com/coq-community/gaia} maintains Grimm's  theory of ordinals as part of coq-community on GitHub. Integration
of the present ordinal theory with Gaia, i.e., relating the different notions of ordinals
and transferring relevant results, would be an interesting project.
